Finiteness and Cardinality in Homotopy Type Theory

### 2020-01-07 paper.tex
corrected closure proofs on cardinal finiteness

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.
\subsubsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

@@

\subsubsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the classification of finiteness predicates mirrors that of
their countable counterparts, and we will prove closure under the Kleene star.

\subsubsection{Practical Uses}
Finally, in section~\ref{practical} we will demonstrate some practical uses of
the finiteness proofs in Cubical Agda.
We will show how to use well-known techniques
\cite{firsovDependentlyTypedProgramming2015, devrieseBrightSideType2011} to
automate proofs of functions with finite (and infinite) domains.
We further show how to automate the synthesis of functions from desired
properties.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.

@@

@@

@@

  We then need to convert a value of type \(\mathbf{Dec}(x \in \mathit{xs})\) to
  \(x \in \mathit{xs}\).
  We use the following to do that:
  \begin{equation}
    \text{recompute} : \mathbf{Dec}(A) \rightarrow \lVert A \rVert \rightarrow A
  \end{equation}
  The function works via case-analysis on \(\mathbf{Dec}\): in the true case, we
  return the proof; in the false case, we can apply the proof of negation under
  the truncation, and then extract \(\lVert \bot \rVert\) to \(\bot\), as
  \(\bot\) is a mere proposition, thus proving the goal via explosion.
\end{proof}
\subsection{Closure}
\begin{rm-lemma}
  Manifest enumerability is closed under \(\Sigma\).
\end{rm-lemma}
\begin{proof}
  For these three closures, the proofs on split enumerability consisted of a
  list manipulation followed by a proof that membership was preserved by the
  list manipulation.
  Because we separate these two concerns, the proofs carry over onto manifest
  enumerability: the support list manipulation stays the same, and the
  cover proofs are performed ``under'' the truncation.
\end{proof}
\section{Cardinal Finiteness} \label{cardinal}
For manifest enumerability, we removed the need for decidable equality: in these
next two finiteness predicates, we remove the need for a total order on the
underlying type.
\begin{rm-definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite, \(\mathcal{C}\), if it has a
  propositionally-truncated proof of bishop finiteness.
  \begin{equation}
    \mathcal{C}(A) = \lVert \mathcal{B}(A) \rVert
  \end{equation}
\end{rm-definition}
\subsection{Closure}
\begin{rm-lemma}
  Cardinal finiteness is closed under \(\Sigma\) and \(\Pi\).
\end{rm-lemma}
\begin{proof}
  All closure functions can be lifted under propositional truncation.
  Therefore, cardinal finiteness has the same closure properties as manifest
  Bishop finiteness.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-theorem} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{rm-theorem}
\begin{proof}
  We will use eliminator~\ref{elim-prop-prop} from definition~\ref{prop-trunc}.
  Manifest Bishop finiteness implies decidable equality already, so our task
  here is to prove that decidable equality itself is a proposition.

>>>>

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.
\subsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

@@

\subsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the classification of finiteness predicates mirrors that of
their countable counterparts, and we will prove closure under the Kleene star.

\subsection{Practical Uses}
Finally, in section~\ref{practical} we will demonstrate some practical uses of
the finiteness proofs in Cubical Agda.
We will show how to use well-known techniques
\cite{firsovDependentlyTypedProgramming2015, devrieseBrightSideType2011} to
automate proofs of functions with finite (and infinite) domains.
We further show how to automate the synthesis of functions from desired
properties.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.

@@

\begin{rm-theorem} \label{split-enum-sigma}

@@

  \begin{equation} \label{split-enum-sigma-clos-sig}

@@

  We then need to convert a value of type \(\mathbf{Dec}(x \in \mathit{xs})\) to
  \(x \in \mathit{xs}\).
  We use the following to do that:
  \begin{equation}
    \text{recompute} : \mathbf{Dec}(A) \rightarrow \lVert A \rVert \rightarrow A
  \end{equation}
  The function works via case-analysis on \(\mathbf{Dec}\): in the true case, we
  return the proof; in the false case, we can apply the proof of negation under
  the truncation, and then extract \(\lVert \bot \rVert\) to \(\bot\), as
  \(\bot\) is a mere proposition, thus proving the goal via explosion.
\end{proof}
\subsection{Closure}
Since we don't have an equivalence with \(\mathbf{Fin}\), we don't get closure
under \(\Pi\).
\begin{rm-lemma}
  Manifest enumerability is closed under \(\Sigma\).
\end{rm-lemma}
\begin{proof}
  This closure proof is almost the same as theorem~\ref{split-enum-sigma}.
  The manipulation of the support lists can be carried over as-is; but the type
  of the cover proof has changed, so it will need to be updated.
  As it happens, the translation is straightforward: we effectively write the
  ``monadic'' version of the old function.
\end{proof}
\section{Cardinal Finiteness} \label{cardinal}
For manifest enumerability, we removed the need for decidable equality: in these
next two finiteness predicates, we remove the need for a total order on the
underlying type.
\begin{rm-definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite, \(\mathcal{C}\), if it has a
  propositionally-truncated proof of bishop finiteness.
  \begin{equation}
    \mathcal{C}(A) = \lVert \mathcal{B}(A) \rVert
  \end{equation}%
\end{rm-definition}%
\subsection{Closure}
The closure properties of cardinal finiteness are effectively the non-dependent
versions of manifest Bishop finiteness.
To see why, consider equation~\ref{split-enum-sigma-clos-sig}.
We can ``lift'' the proof (as a binary function) under a propositional
truncation, giving us equation~\ref{sigma-proof-lifted}, but that doesn't give us
the desired closure proof (equation~\ref{sigma-proof-desired}).
\begin{multicols}{2}
  \noindent
  \begin{equation} \label{sigma-proof-lifted}
    \frac{
      \lVert \mathcal{B}(A) \rVert \; \; \; \lVert \Pi (x : A) , \mathcal{B}(U(x)) \rVert
    }{
      \lVert \mathcal{B}(\Sigma(x : A) , U(x)) \rVert
    }
  \end{equation}
  \begin{equation} \label{sigma-proof-desired}
    \frac{
      \lVert \mathcal{B}(A) \rVert \; \; \; \Pi (x : A) , \lVert \mathcal{B}(U(x))  \rVert
    }{
      \lVert \mathcal{B}(\Sigma(x : A) , U(x)) \rVert
    }
  \end{equation}
\end{multicols}%
To remedy the mismatch we would need a function of the type:
\begin{equation}
  \left( \Pi(x : A) , \lVert \mathcal{B}(U(x)) \rVert \right) \rightarrow
  \lVert \Pi(x : A) , \mathcal{B}(U(x)) \rVert
\end{equation}
Unfortunately, this equation in particular is a form of the axiom of choice
\cite[equation 3.8.3]{hottbook}.
This leaves us with closure under only the non-dependent operations.
\begin{rm-lemma}
  Cardinal finiteness is closed under \(\times\), \(+\), and \(\rightarrow\).
\end{rm-lemma}
\begin{proof}
  All of these closure proofs can be lifted directly from their corresponding
  proofs on manifest Bishop finiteness.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-theorem} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{rm-theorem}
\begin{proof}
  We will use eliminator~\ref{elim-prop-prop} from definition~\ref{prop-trunc}.
  Manifest Bishop finiteness implies decidable equality already, so our task
  here is to prove that decidable equality itself is a proposition.
